---
title: "CS 422"
author: "Kamen Petkov, Illinois Institute of Technology"
output:
  html_document:
    toc: yes
    df_print: paged
  html_notebook:
    toc: yes
    toc_float: yes
---

### Libraries
```{r}
library(dplyr)
library(psych)
library(cluster)
library(factoextra)
library(dbscan)
library(fpc)

setwd("C:/Users/ACER/Desktop/Desktop/IIT_Stuff/CS 422/myHws/HW8")
```

### (a) Data Cleanup
```{r}
# Data that we're going to use
df <- read.table("file19.txt", skip = 20, header = TRUE)

# Show part of data 
head(df)

# Find correlation between the attribute
cor(df[,-1])

# (i)
df$c = NULL
df$p = NULL
df$m = NULL

summary(df)

# (ii) Standardize the data
df.scale <- scale(df[,-1])
summary(df.scale)


# (iii)
write.csv(df,"file19.csv")
```
#### (i) Think of what attributes, if any, you may want to omit from the dataset when you do the clustering. Indicate all of the attributes you removed before doing the clustering.
#### Drop c,p,m since they are strongly correlated with C,P,M - we don't need both.

#### (ii) Does the data need to be standardized?
#### In this case standardization might not be necessary since the difference between min and max is not too big. However, for the sake of making the work easier for the computer (algorithm is easier to use), I will scale.


### B) Clustering
```{r}
fviz_nbclust(df.scale[, -1], kmeans, method = "wss")
fviz_nbclust(df.scale[, -1], kmeans, method = "silhouette")

# (ii) Once you have determined the number of clusters, run k-means clustering on the dataset to create that many
# clusters. Plot the clusters using fviz_cluster().

clst <- kmeans(df.scale[, -1], centers = 8, nstart = 25)
fviz_cluster(clst, df.scale[, -1], main = "Cluster")

# (iii) How many observations are in each cluster?
clst$size


# (iv) What is the total SSE of the clusters?
clst$totss
clst$tot.withinss

# Total SSE of the clusters = 260
# Total within-clusters = 17.57414 

# (v) What is the SSE of each cluster?
clst$withinss

# (vi) Perform an analysis of each cluster to determine how the mammals are grouped in each cluster, and whether
# that makes sense?

df[which(clst$cluster == 1),]
df[which(clst$cluster == 2),]
df[which(clst$cluster == 3),]
df[which(clst$cluster == 4),]
df[which(clst$cluster == 5),]
df[which(clst$cluster == 6),]
df[which(clst$cluster == 7),]
df[which(clst$cluster == 8),]
```
#### (i) I chose to have 8 clusters because from the wss graph, we can notice that after 8, there isn't a big change in the numbers. Perhaps, ideally we would use 10.

#### (iii) Number of observations in each cluster: 1, 8, 7, 11, 6, 8, 13, 12

#### (iv) Total SSE of the clusters = 260; Total within-clusters = 17.57414 -> I believe we are looking for this one.

#### (v) SSE of each cluster: 0.000000e+00 5.908626e-01 3.882675e-31 1.841650e+00 5.083201e+00 1.471193e+00 1.869983e+00 6.717249e+00


#### (vi) Perform an analysis of each cluster to determine how the mammals are grouped in each cluster, and whether that makes sense?

#### Cluster 1 makes sense since there is only 1 member in it - Armadillo.
#### Cluster 2 makes sense. The attributes remain the same in all observations. Moreover, all of the animals are rodents in this cluster
#### In cluster 3 all attributes remain the same in the observations. They are all from cervidae family.
#### Cluster 4 makes sense. All attributes except for attribute P remain the same in the observations. The animals are rodents.
#### Cluster 5 contains attributes that remain more or less the same in the observations. The cluster definitely makes sense when viewed from the perspective that all of these sea animals look similar to each other (especially their tusks).
#### Even though in cluster 6 the attributes are the same for the most part in the observations, the cluster doesn't make sense to me when it comes to finding similarities between the animals in it.
#### Cluster 7 makes sense. Attributes remain the same for the most part. The animals in it are similar. They have many cat characteristics. 
#### Cluster 8 doesn't make sense to me.